What We Eval in the Shadows

### 2021-04-16 paper/main.tex
Grammarly's proofreading of the expression of eval

\begin{wrapfigure}{r}{5cm} \hspace*{-12mm}
  \centering
  \includegraphics[width=72mm]{pkgs-eval-callsites-hist.pdf} \caption{CRAN
  \eval call sites}%
  \label{fig:pkgs-eval-callsites-hist}
\end{wrapfigure}
%
This section focuses on CRAN packages and reports statistics about the usage of
\eval. We use the word \emph{site} to an occurrence of a call site to the \eval
function in the source code, and \emph{call} to denote an observed invocation of
the \eval function.

@@

\medskip\noindent\EE{$min(e)=\c{V}$} Expressions that represent values are
frequently passed to \eval -- they occur in 17\% of sites. The majority of
those, \packageValOneNodePercent, are inline constants (integer or double
vectors). The rest trivially evaluate to a value, \eg~\c{1+1}.\footnote{True as
long as base functions such as \c{+} are not redefined. They typically aren't,
but this is a limitation of this categorization.} In our corpus,
\packageNbCallSitesUniqueActualValue call sites only ever see a simple value.
Manual inspection reveals cases such as the following
\begin{lstlisting}
 f1 <- eval(paste("A~",paste(paste(names(X[,-1])),collapse="+")))
\end{lstlisting}
where the argument to \eval is a string which \eval simply returns. From twenty
randomly selected \evals, \packageUsefulValueEvalPercent of them  need % There are a majority which needs eval
\eval for some inputs. Either we are dealing with a value that needs to be
constructed dynamically or the value is a default case that sometimes is
replaced by a more interesting expression. This form is usually evaluated in few
interpreter steps, in fact the median is only
\packageMinimizedmedianoperationscRnd. The environment in which they evaluate is
mostly irrelevant (unless a built-in operator is redefined).

\medskip\noindent\EE{$min(e)=\c{X}$} Variable lookups are the most common form,
they are found in 28\% of the sites. This form includes simple variable reads,
\eg~\c{x}, those are \packageNbSymbolVarSitePercent of \c X. The form also
subsumes \c{V}, so it includes a mixture of arithmetic expressions,
\eg~\c{x+y+1}. The operations allowed are limited to built-in arithmetics. It is
noteworthy that, while most \c{X}\!s evaluate in a single step, the variable can
be bound to a promise, and accessing it may trigger evaluation of that promise,
thus resulting in an arbitrary amount of computation. The median number of
interpreted operations is \packageMinimizedmedianoperationsaRnd, suggesting that
it is not the common case. Lookups are often evaluated in constructed
environments, as few as \packageMinimizedpercentparentframesc of these
expressions are evaluated in a function environment.

\medskip\noindent\EE{$min(e)=\c{\$}$} This form extends \c X to include lookup
with the dollar operator, \eg~\c{x\$f}, and vector indexing, \eg~\c{x[42]} or
\c{x[[24]]}. As with \c X we allow arithmetics and values in this form. Lookup
occurs in 6\% of the sites. The interpreter evaluates
\packageMinimizedmedianoperationsgRnd operations on average; the minimum is 3
operations. This is typically used in a function environment,
\packageMinimizedpercentparentframese of the time to be precise.

\medskip\noindent\EE{$min(e)=$~\c{<-}} This form includes both assignment
operators, the direct assignment {\tt <-}, and assignment to the parent
environment {\tt <\,\!<-}, the \c{assign} function and the \c{\$} form. Assignments occur in 5\% of the
sites. They represent the most obvious source of side-effects. The median number
of operations is \packageMinimizedmedianoperationsiRnd; the minimum is 3.

\medskip\noindent\EE{$min(e)=\c{F()}$} This form captures simple function calls,
\eg~\c{f(2)}, with neither variables nor assignments. Or, more specifically it
allows for variables in the function position but not in arguments. Usually,
looking up function names does not trigger computation, but that is not a given
if the function is returned by a promise or if the function name is shadowed by
a promise containing a value, then computation will occur. This form occurs in
6\% of sites and typically does not perform much work in the interpreter.

\medskip\noindent\framebox{$min(e)=\c{F(X)}$}~\EE{$min(e)=\c{F(F(X))}$} These
forms allow for function calls whose arguments may include variable references.
The latter allows nested calls. Assignments are excluded from this form. They
occur in, respectively, 14\% and 20\% of the sites in the corpus. Together they
are the most frequent forms. The median numbers of interpreter steps are,
respectively, \packageMinimizedmedianoperationsdRnd and
\packageMinimizedmedianoperationsbRnd. The \packageMinimizedpercentparentframesj
of these expressions run in function environments.

\medskip\noindent\EE{$min(e)=\c{FUN}$} This form captures expressions that define functions, \eg~\c{function(x) x+1}, and do nothing else. \c{FUN} occurs in
\packageFunctionDefinitionSitesPercent of sites. Evaluating a function
definition is done in 2 interpreter step; the data does not record the work
performed by the interpreter when the generated functions are eventually run. In
addition to \c{FUN}, \packageGeneralizedFunctionDefinitionSitesPercent of sites
have function definitions nested in other expressions, this gives an idea of the
use of higher-order functions.

\medskip\noindent\EE{$min(e)=\c{BLOCK}$} This form captures multi-statement
code blocks, which occur in only 4\% of sites. These are larger expressions, we
do not inspect contents of the blocks. The median number of executed operations
is \packageMinimizedmedianoperationsjRnd. They typically run in function
environments.

\medskip\noindent\EE{$min(e)=\c{model.frame}$} The \c{model.frame} function
returns a dataframe that is the result of fitting the model described in a given
formula. This form subsumes \c{F(F(X))}, \c{FUN} and assignments. It is the
single most popular function invoked from \eval, it occurs in 7\% of the sites.
Each call does quite a lot of work with a 2K instructions median.

\paragraph{Consistency} It is interesting to consider how many different
forms any given site sees. The more forms, the harder it will be to characterize
the behavior of the program at that site. Luckily, \packageNbOneMinimizedPercent
of sites only see a single form.  In other words, a JIT compiler could correctly predict the \emph{shape} of an expression by looking just at the first call in \packageNbOneMinimizedPercent of the sites.  There are few sites that are highly polymorphic (8 or more different forms), these include the pipe operator of the \c{magrittr} package which is used to compose functions.

\paragraph{Discussion} The variety of uses of \eval is evidenced by the number
of different forms observed in CRAN. In comparison, JavaScript \eval usage was
simpler and more predictable as reported by \citet{oopsla12b}. Nevertheless,
simple forms dominate and there are many cases where \eval could be replaced
by less powerful constructs.

>>>>

\begin{wrapfigure}{r}{5cm} \hspace*{-12mm}
  \centering
  \includegraphics[width=72mm]{pkgs-eval-callsites-hist.pdf} \caption{CRAN
  \eval call sites}%
  \label{fig:pkgs-eval-callsites-hist}
\end{wrapfigure}
%
This section focuses on CRAN packages and reports statistics about the usage of
\eval. We use the word \emph{site} to an occurrence of a call site to the \eval
function in the source code and \emph{call} to denote an observed invocation of
the \eval function.

@@

\medskip\noindent\EE{$min(e)=\c{V}$} Expressions that represent values are
frequently passed to \eval -- they occur in 17\% of sites. The majority of
those, \packageValOneNodePercent, are inline constants (integer or double
vectors). The rest trivially evaluate to a value, \eg~\c{1+1}.\footnote{True as
long as base functions such as \c{+} are not redefined. They typically are not,
but this is a limitation of this categorization.} In our corpus,
\packageNbCallSitesUniqueActualValue call sites only ever see a simple value.
Manual inspection reveals cases such as the following
\begin{lstlisting}
 f1 <- eval(paste("A~",paste(paste(names(X[,-1])),collapse="+")))
\end{lstlisting}
where the argument to \eval is a string which \eval simply returns. Of twenty
randomly selected \evals, \packageUsefulValueEvalPercent of them  need % There are a majority which needs eval
\eval for some inputs. Either we are dealing with a value that needs to be
constructed dynamically, or the value is a default case that sometimes is
replaced by a more interesting expression. This form is usually evaluated in few
interpreter steps; in fact, the median is only
\packageMinimizedmedianoperationscRnd. The environment in which they evaluate is
mostly irrelevant (unless a built-in operator is redefined).

\medskip\noindent\EE{$min(e)=\c{X}$} Variable lookups are the most common form; they are found in 28\% of the sites. This form includes simple variable reads,
\eg~\c{x}, which are \packageNbSymbolVarSitePercent of \c X. The form also
subsumes \c{V}, so it includes a mixture of arithmetic expressions,
\eg~\c{x+y+1}. The operations allowed are limited to built-in arithmetics. It is
noteworthy that, while most \c{X}\!s evaluate in a single step, the variable can
be bound to a promise, and accessing it may trigger evaluation of that promise,
thus resulting in an arbitrary amount of computation. The median number of
interpreted operations is \packageMinimizedmedianoperationsaRnd, suggesting that
it is not the common case. Lookups are often evaluated in constructed
environments; as few as \packageMinimizedpercentparentframesc of these
expressions are evaluated in a function environment.

\medskip\noindent\EE{$min(e)=\c{\$}$} This form extends \c X to include lookup
with the dollar operator, \eg~\c{x\$f}, and vector indexing, \eg~\c{x[42]} or
\c{x[[24]]}. As with \c X, we allow arithmetics and values in this form. Lookup
occurs in 6\% of the sites. The interpreter evaluates
\packageMinimizedmedianoperationsgRnd operations on average; the minimum is 3
operations. This is typically used in a function environment,
\packageMinimizedpercentparentframese of the time to be precise.

\medskip\noindent\EE{$min(e)=$~\c{<-}} This form includes both assignment
operators, the direct assignment {\tt <-}, and assignment to the parent
environment {\tt <\,\!<-}, the \c{assign} function and the \c{\$} form. Assignments occur in 5\% of the
sites. They represent the most obvious source of side effects. The median number
of operations is \packageMinimizedmedianoperationsiRnd; the minimum is 3.

\medskip\noindent\EE{$min(e)=\c{F()}$} This form captures simple function calls,
\eg~\c{f(2)}, with neither variables nor assignments. More specifically, it
allows for variables in the function position but not in arguments. Usually,
looking up function names does not trigger computation, but that is not a given
if the function is returned by a promise or if the function name is shadowed by
a promise containing a value, then the computation will occur. This form occurs in
6\% of sites and typically does not perform much work in the interpreter.

\medskip\noindent\framebox{$min(e)=\c{F(X)}$}~\EE{$min(e)=\c{F(F(X))}$} These
forms allow for function calls whose arguments may include variable references.
The latter allows nested calls. Assignments are excluded from this form. They
occur in, respectively, 14\% and 20\% of the sites in the corpus. Together they
are the most frequent forms. The median numbers of interpreter steps are,
respectively, \packageMinimizedmedianoperationsdRnd and
\packageMinimizedmedianoperationsbRnd. Also,  \packageMinimizedpercentparentframesj
of these expressions run in function environments.

\medskip\noindent\EE{$min(e)=\c{FUN}$} This form captures expressions that define functions, \eg~\c{function(x) x+1}, and do nothing else. \c{FUN} occurs in
\packageFunctionDefinitionSitesPercent of sites. Evaluating a function
definition is done in 2 interpreter steps; the data does not record the work
performed by the interpreter when the generated functions are eventually run. In
addition to \c{FUN}, \packageGeneralizedFunctionDefinitionSitesPercent of sites
have function definitions nested in other expressions. This gives an idea of the
use of higher-order functions.

\medskip\noindent\EE{$min(e)=\c{BLOCK}$} This form captures multi-statement
code blocks, which occur in only 4\% of sites. These are huge expressions; we
do not inspect the contents of the blocks. The median number of executed operations
is \packageMinimizedmedianoperationsjRnd. They typically run in function
environments.

\medskip\noindent\EE{$min(e)=\c{model.frame}$} The \c{model.frame} function
returns a dataframe resulting from f fitting the model described in a given
formula. This form subsumes \c{F(F(X))}, \c{FUN}, and assignments. It is the
single most popular function invoked from \eval; it occurs in 7\% of the sites.
Each call does quite a lot of work with a 2K instructions median.

\paragraph{Consistency} It is interesting to consider how many different
forms any given site sees. The more forms, the harder it will be to characterize
the program's behavior at that site. Luckily, \packageNbOneMinimizedPercent
of sites only see a single form.  In other words, a JIT compiler could correctly predict the \emph{shape} of an expression by looking just at the first call in \packageNbOneMinimizedPercent of the sites.  Few sites are highly polymorphic (8 or more different forms). These include the pipe operator of the \c{magrittr} package, which is used to compose functions.

\paragraph{Discussion} The variety of uses of \eval is evidenced by the number
of different forms observed in CRAN. In comparison, JavaScript \eval usage was more straightforward and more predictable as reported by \citet{oopsla12b}. Nevertheless, simple forms dominate, and there are many cases where \eval could be replaced by less powerful constructs.
